No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-05-29 paper/GEB Submission/manuscript_GEB.Rmd
line breaks

Broad-scale trends in functional diversity are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021]. While press disturbances can also lead to loss of functional volume as species cannot adapt rapidly to new conditions, it can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.

`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023; @larson2021].
`r colorize("While loss of functional diversityis frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020]. `r colorize("Broad-scale studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016]. `r colorize("We have stronger expectations for changes in the prevalence of some individual traits. For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018]. `r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].

@@

`r colorize("Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact. For example, we found no trend for mean body mass in bird or mammal communities. In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size. This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts", "blue") `[@hantak2021]`r colorize(", or current shifts are happening at an intraspecific level not captured by our data. Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].

>>>>

Broad-scale trends in functional diversity are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
While press disturbances can also lead to loss of functional volume as species fail to adapt to new conditions, it can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.

`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023; @larson2021].
`r colorize("While loss of functional diversity is frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020].
`r colorize("Broad-scale studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016].
`r colorize("We have stronger expectations for changes in the prevalence of some individual traits. For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
`r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].
`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].

@@

`r colorize("Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact. For example, we found no trend for mean body mass in bird or mammal communities. In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size. This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts", "blue")`[@hantak2021]`r colorize(", or current shifts are happening at an intraspecific level not captured by our data. Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
